---
date: 2023-11-27
lesson: SpatAn1
topic: Spatial Joins
index: 2
execute:
  echo: false   # set to true to show sample solution
  output: false # set to true to show sample solution
code-fold: true
code-summary: "Sample Solution"
knitr:
  opts_chunk:
    collapse: true  
---

# SpatAn1 1: Exercise B {#sec-rauman1b}

For the upcoming exercise, we will work with the *gruental.gpkg*. data set. Import it into R. We also need the following libraries:

```{r}
#| echo: true
#| code-fold: false

library("dplyr")
library("sf")
library("ggplot2")
```

```{r}
#| echo: false
#| output: false

read_sf("datasets/rauman/gruental.gpkg")
```

### Task 1: Geopackage "Layers"

You may have noticed the following warning message when importing the *gruental.gpkg* geopackage:

```
Warning message:
In evalq((function (..., call. = TRUE, immediate. = FALSE, noBreaks. = FALSE,  :
  automatically selected the first layer in a data source containing more than one.
```

This warning message indicates that the geopackage `gruental.gpkg` has several layers (rep. records) and only the first layer has been imported. Use the `st_layers` command to find out the layer names and then use them in `st_read` (as argument `layer =`) to import the layers individually and store them in variables (e.g., such as in the variables `wiesen` and `baeume`).

```{r}
st_layers("datasets/rauman/gruental.gpkg")

meadows <- read_sf("datasets/rauman/gruental.gpkg", "wiesen")
trees <- read_sf("datasets/rauman/gruental.gpkg", "baeume")
```

### Task 2: Understanding the data sets

Take some time to explore the two datasets. Use the visualisation options of `ggplot` (especially `geom_sf`). You can superimpose multiple `geom_sf` to represent multiple records at the same time.

```{r}
#| output: true
#| label: fig-ggplot-ex2
#| layout-ncol: 2
#| column: page
#| fig.cap:
#| - Meadow areas are shown in different colours depending on type 
#| - Trees are shown in different colours depending on type

ggplot(meadows) +
  geom_sf(aes(fill = flaechen_typ)) +
  geom_sf(data = trees) +
  theme_void()

ggplot(meadows) +
  geom_sf() +
  geom_sf(data = trees, aes(colour = baum_typ)) +
  theme_void()
```

### Task 3:  Spatial join with points {#sec-spatial-join-points}

We now want to know whether each *tree* is in a meadow or not. To do this, we use the GIS technique *spatial join* as described in the lecture. Using `sf`, we can perform spatial joins with `st_join`. There are only `left` and `innerjoins` (see PrePro 1 & 2). The "links" points must be listed first, since we want to attach attributes *to* the points.

Note that the output has a new column: `flaechen_typ`. This is empty (`NA`) if the corresponding tree is not in a meadow. How many trees are in a meadow and how many are not?

```{r}
trees_join <- st_join(trees, meadows)

number_in_meadow <- sum(!is.na(trees_join$flaechen_typ))
number_not_in_meadow <- sum(is.na(trees_join$flaechen_typ))
```
