---
date: 2024-04-02
lesson: stat1
topic: Statistic 1
index: 1
execute:
  echo: true # set to true to show sample solution 
  output: true # set to true to show sample solution
code-fold: true
code-summary: "Sample Solution"
knitr:
  opts_chunk:
    collapse: true  
---

# Statistics 1: The basics of Statistics

Exercise 1  
Use the data in WAGE1.RAW for this exercise.  
1) Find the average education level in the sample. What are the lowest and highest years of education?  
2) Find the average hourly wage in the sample. Does it seem high or low?  
3) The wage data are reported in 1976 dollars. Using the Economic Report of the President (2011 or later), obtain and report the Consumer Price Index (CPI) for the years 1976 and 2010.  
4) Use the CPI values from 3) to find the average hourly wage in 2010 dollars. Now does the average hourly wage seem reasonable?  
5) How many women are in the sample? How many non-women?

```{r}
#| eval: false
library('wooldridge')
data('wage1')
?wage1

#1)
mean(wage1$educ)
min(wage1$educ)
max(wage1$educ)

#2)
mean(wage1$wage)
#This seems low

#3)
#CPI in 1976 is 56.9
#CPI in 2010 is 218.056

#4)
mean(wage1$wage)*218.056/56.9

#5)
nrow(wage1[wage1$female == 1, ])
nrow(wage1[wage1$female == 0, ])

```

Exercise 2  
Use the data in BWGHT.RAW to answer this question.  
1) How many women are in the sample, and how many report smoking during pregnancy?  
2) What is the average number of cigarettes smoked per day? Is the average a good measure of the “typical” woman in this case? Explain.  
3) Among women who smoked during pregnancy, what is the average number of cigarettes smoked per day? How does this compare with your answer from 2), and why?  
4) Find the average of $fatheduc$ in the sample. Why are only 1,192 observations used to compute this average?  
5) Report the average family income and its standard deviation in dollars.  

```{r}
#| eval: false
library('wooldridge')
library('dplyr')
data('bwght')
?bwght

#1)
nrow(bwght[is.na(bwght$mothereduc), ])
#there is no missing information so we have 1388 women
1388-nrow(bwght[(bwght$cigs==0), ])

#2)
mean(bwght$cigs)
#No, because there is a lot of non-smokers

#3)
bwght %>% filter(cigs>0) %>% summarise(mean(cigs))

#4)
mean(bwght$fatheduc, na.rm = TRUE)

#5)
mean(bwght$faminc)
sd(bwght$faminc)

```

Exercise 3  
The data in MEAP01.RAW are for the state of Michigan in the year 2001. Use these data to answer the following questions.  
1) Find the largest and smallest values of $math4$. Does the range make sense? Explain.  
2) How many schools have a perfect pass rate on the math test? What percentage is this of the total sample?  
3) How many schools have math pass rates of exactly 50%?  
4) Compare the average pass rates for the math and reading scores. Which test is harder to pass?  
5) Find the correlation between $math4$ and $read4$. What do you conclude?  
6) The variable $exppp$ is expenditure per pupil. Find the average of $exppp$ along with its standard deviation. Would you say there is wide variation in per pupil spending?  
7) Suppose School A spends \$6,000 per student and School B spends \$5,500 per student. By what percentage does School A’s spending exceed School B’s?  

```{r}
#| eval: false
library('wooldridge')
data('meap01')
?meap01

#1)
min(meap01$math4)
max(meap01$math4)

#2)
nrow(meap01[(meap01$math4==100), ])
nrow(meap01[(meap01$math4==100), ])/1823*100

#3)
nrow(meap01[(meap01$math4==50), ])

#4)
mean(meap01$math4)
mean(meap01$read4)

#5)
corr=lm(math4 ~ read4, data = meap01)
summary(corr)

#6)
mean(meap01$exppp)
sd(meap01$exppp)

#7)
(6000-5500)/5500*100

```

Exercise 4  
The data in JTRAIN2.RAW come from a job training experiment conducted for low-income men during 1976–1977.  
1) Use the indicator variable $train$ to determine the fraction of men receiving job training.  
2) The variable $re78$ is earnings from 1978, measured in thousands of 1982 dollars. Find the averages of $re78$ for the sample of men receiving job training and the sample not receiving job training. Is the difference economically large?  
3) The variable $unem78$ is an indicator of whether a man is unemployed or not in 1978. What fraction of the men who received job training are unemployed? What about for men who did not receive job training? Comment on the difference.  
4) From questions 2) and 3), does it appear that the job training program was effective? What would make our conclusions more convincing?  


```{r}
#| eval: false
library('wooldridge')
data('jtrain2')
?jtrain2

#1)
nrow(jtrain2[(jtrain2$train==1), ])/445

#2)
jtrain2 %>% filter(train==0) %>% summarise(mean(re78))
jtrain2 %>% filter(train==1) %>% summarise(mean(re78))

#3)
(nrow(jtrain2[(jtrain2$train==1), ]) - (jtrain2 %>% filter(train==1) %>% summarise(sum(unem78))))/nrow(jtrain2[(jtrain2$train==1), ])

(nrow(jtrain2[(jtrain2$train==0), ]) - (jtrain2 %>% filter(train==0) %>% summarise(sum(unem78))))/nrow(jtrain2[(jtrain2$train==0), ])

#4)
#No
#To make a linear regression to find if there is evidence on the fact that the training does not have any effects on the employement rate

```

Exercise 5  
The data in FERTIL2.DTA were collected on women living in the Republic of Botswana in 1988. The variable children refers to the number of living children. The variable electric is a binary indicator equal to one if the woman’s home has electricity, and zero if not.  
1) Find the smallest and largest values of $children$ in the sample. What is the average of $children$?  
2) What percentage of women have electricity in the home?  
3) Compute the average of $children$ for those without electricity and do the same for those with electricity. Comment on what you find.  
4) From question 3), can you infer that having electricity “causes” women to have fewer children? Explain.  

```{r}
#| eval: false
library('wooldridge')
data('fertil2')
?fertil2

#1)
min(fertil2$children)
max(fertil2$children)
mean(fertil2$children)

#2)
nrow(fertil2[(fertil2$electric==1), ])/4361*100

#3)
fertil2 %>% filter(electric==0) %>% summarise(mean(children))
fertil2 %>% filter(electric==1) %>% summarise(mean(children))

#4)
#No

```
